---
title: "Statistical Inference in Nutshell"
description: "Visual and intuitive explanation of statistical inference."
categories: ["Statistics"]
author: "Jiho Kim"
date: 2024-05-03
html-math-method: katex
code-annotations: below
---

## Single Proportion

### Confidence Interval

If the population follows a normal distribution or if the sample size $n$ is sufficiently large so that $n \hat{p} \geq 10$ and $n (1 - \hat{p}) \geq 10$ for the sample proportion $\hat{p}$, then a confidence interval for the population proportion $p$ can be calculated as: $$\hat{p} \pm z^* \sqrt{\dfrac{\hat{p} (1 - \hat{p})}{n}}$$ where $z^*$ is chosen such that $P(-z^* < Z < z^*) = 1 - \alpha$ for a given significance level $\alpha$.

#### Example

Suppose we are given a sample size of $n = 400$, a sample proportion of $\hat{p} = 0.84$, and a significance level of $\alpha = 0.05$ (or a $1 - \alpha = 0.95$ confidence level). 

Without knowing whether the population follows a normal distribution, we verify that both $n \hat{p} = (400)(0.84) = 336 \geq 10$ and $n (1 - \hat{p}) = (400)(0.16) = 64 \geq 10$ are satisfied. Therefore, we can proceed with calculating the confidence interval:

```{r}
n <- 400
p_hat <- 0.84
alpha <- 0.05

z_star <- qnorm(1 - (alpha / 2))                        # <1>
standard_error <- sqrt((p_hat * (1 - p_hat)) / n)
interval <- p_hat + c(-1, 1) * z_star * standard_error  # <2>

cat("(", round(interval[1], 4), ", ", round(interval[2], 4), ")", sep = "")
```
1. Divide alpha by 2 to determine the area for one tail, then subtract it from 1 to obtain the area up to the right tail. Then find the z-score that corresponds to this area using the `qnorm` function.
2. Multiplying the critical value by `c(-1, 1)` is just a lazy yet convenient way to calculate the lower and upper bounds of the confidence interval simultaneously.

<!-- ```{r}
n <- 400
p_hat <- 0.84
conf_level <- 0.95

z_star <- qnorm((conf_level + 1) / 2)                   # <1>
standard_error <- sqrt((p_hat * (1 - p_hat)) / n)
interval <- p_hat + c(-1, 1) * z_star * standard_error

cat("(", round(interval[1], 4), ", ", round(interval[2], 4), ")", sep = "")
```
1. The `qnorm` function takes the area up to the right tail as an argument, so we add 1 to the confidence level and divide it by 2 to get the area up to the right tail. -->

<!-- Here is how to interpret the confidence interval:

> We are not certain whether the true population proportion is within the constructed interval; however, the procedure used to construct the interval has a 95% success rate of capturing the true population proportion. -->

### Determination of Sample Size

To determine the sample size $n$ required to estimate the population proportion $p$ with a desired margin of error $ME$ and a significance level $\alpha$, we can use the following formula: $$n = \lceil\bigg(\dfrac{z^*}{ME}\bigg)^2 \tilde{p} (1 - \tilde{p})\rceil$$ where $z^*$ is chosen such that $P(-z^* < Z < z^*) = 1 - \alpha$ for a given significance level $\alpha$ and $\tilde{p}$ is an estimate of the population proportion $p$ (use $\tilde{p} = 0.5$ if no estimate is available).

### Hypothesis Testing

To test the null hypothesis $H_0: p = p_0$ against the alternative hypothesis $H_1: p \neq p_0$ (or a one-tailed alternative), we use the following test statistic: $$z = \dfrac{\hat{p} - p_0}{\sqrt{\dfrac{p_0 (1 - p_0)}{n}}}$$ where $n$ is the sample size and $\hat{p}$ is the sample proportion, provided that the population follows a normal distribution or that the sample size $n$ is sufficiently large so that $n p_0 \geq 10$ and $n (1 - p_0) \geq 10$.

There are two approaches to hypothesis testing: critical value approach and p-value approach.

#### Critical Value Approach

Suppose we test the null hypothesis $H_0: p = 0.8$ against the alternative hypothesis $H_1: p \neq 0.8$ with a significance level of $\alpha = 0.05$. Then we can calculate the critical value as follows:

```{r}
alpha <- 0.05
z_star <- qnorm(1 - (alpha / 2))
z_star
```

We collect a sample of size $n = 400$ with a sample proportion of $\hat{p} = 0.84$. Without knowing whether the population follows a normal distribution, we verify that both $np_0 = (400)(0.8) = 320 \geq 10$ and $n(1 - p_0) = (400)(0.2) = 80 \geq 10$ are satisfied. Therefore, we can proceed with calculating the test statistic:

```{r}
n <- 400
p_hat <- 0.84
p_0 <- 0.8

z <- (p_hat - p_0) / sqrt((p_0 * (1 - p_0)) / n)
z
```

If the absolute value of the test statistic $|z|$ is greater than the critical value $z^*$, we reject the null hypothesis. Otherwise, we fail to reject the null hypothesis.

```{r}
abs(z) > z_star
```

Therefore, we reject the null hypothesis.

#### P-Value Approach

Suppose we test the null hypothesis $H_0: p = 0.8$ against the alternative hypothesis $H_1: p \neq 0.8$ with a significance level of $\alpha = 0.05$.

We collect a sample of size $n = 400$ with a sample proportion of $\hat{p} = 0.84$. Without knowing whether the population follows a normal distribution, we verify that both $np_0 = (400)(0.8) = 320 \geq 10$ and $n(1 - p_0) = (400)(0.2) = 80 \geq 10$ are satisfied. Therefore, we can proceed with calculating the test statistic:

```{r}
n <- 400
p_hat <- 0.84
p_0 <- 0.8

z <- (p_hat - p_0) / sqrt((p_0 * (1 - p_0)) / n)
z
```

The p-value is the probability of observing a test statistic as extreme as the one calculated from the collected sample, assuming that the null hypothesis is true. The p-value can be calculated as follows:

```{r}
p_value <- 2 * pnorm(-abs(z))
p_value
```

If the p-value is less than $\alpha$, we reject the null hypothesis. Otherwise, we fail to reject the null hypothesis.

```{r}
alpha <- 0.05
p_value < alpha
```

Therefore, we reject the null hypothesis.

## Difference in Proportions

### Confidence Interval

If the population follows a normal distribution or if the sample sizes $n_1$ and $n_2$ are sufficiently large so that $n_1 \hat{p}_1 \geq 10$ and $n_1 (1 - \hat{p}_1) \geq 10$ and that $n_2 \hat{p}_2 \geq 10$ and $n_2 (1 - \hat{p}_2) \geq 10$ for the sample proportions $\hat{p}_1$ and $\hat{p}_2$, then a confidence interval for the difference in population proportions $p_1 - p_2$ can be calculated as: $$(\hat{p}_1 - \hat{p}_2) \pm z^* \sqrt{\dfrac{\hat{p}_1 (1 - \hat{p}_1)}{n_1} + \dfrac{\hat{p}_2 (1 - \hat{p}_2)}{n_2}}$$ where $z^*$ is chosen such that $P(-z^* < Z < z^*) = 1 - \alpha$ for a given significance level $\alpha$.

#### Example

Suppose we are given sample sizes of $n_1 = 400$ and $n_2 = 600$, sample proportions of $\hat{p}_1 = 0.84$ and $\hat{p}_2 = 0.78$, and a significance level of $\alpha = 0.05$ (or a $1 - \alpha = 0.95$ confidence level). 

Without knowing whether the populations follow a normal distribution, we verify that $n_1 \hat{p}_1 = (400)(0.84) = 336 \geq 10$, $n_1 (1 - \hat{p}_1) = (400)(0.16) = 64 \geq 10$, $n_2 \hat{p}_2 = (600)(0.78) = 468 \geq 10$, and $n_2 (1 - \hat{p}_2) = (600)(0.22) = 132 \geq 10$ are satisfied. Therefore, we can proceed with calculating the confidence interval:

```{r}
n1 <- 400
n2 <- 600
p1_hat <- 0.84
p2_hat <- 0.78
alpha <- 0.05

z_star <- qnorm(1 - (alpha / 2))                                                            # <1>
standard_error <- sqrt(
    ((p1_hat * (1 - p1_hat)) / n1) + ((p2_hat * (1 - p2_hat)) / n2)
)
interval <- (p1_hat - p2_hat) + c(-1, 1) * z_star * standard_error                          # <2>

cat("(", round(interval[1], 4), ", ", round(interval[2], 4), ")", sep = "")
```
1. Divide alpha by 2 to determine the area for one tail, then subtract it from 1 to obtain the area up to the right tail. Then find the z-score that corresponds to this area using the `qnorm` function.
2. Multiplying the critical value by `c(-1, 1)` is just a lazy yet convenient way to calculate the lower and upper bounds of the confidence interval simultaneously.

## Chi-Square Goodness of Fit
## Chi-Square Test for Association

## Single Mean

## Difference in Means
## Paired Difference in Means
## Analysis of Variance (ANOVA)

## Correlation, Simple Regression